\subsection{Exceptions Encoding}

There are better ways of storing the exceptions than in the raw sequential
format of vbe21.
As we have seen there are usually 5 to 6 exceptions per read meaning that
storing the number of exceptions using two bytes, which has a range of 0 to 65535,
for all reads is wasteful.
In particular, the number of exceptions per read ranges from 0 to 3616 and 99\%
of the reads have between 0 and 42 exceptions. This is much smaller than 65535.
See Table \ref{tab:ex} for an statistical overview of the zig-zag delta
exceptions. Its distribution, shown in Figure \ref{fig:nex-hist}, could be
nicely fitted by an exponential distribution. In fact, around 20\% of the reads
have no exceptions at all. There is clearly wasted space being used.

\input{plots/ex-tab}
\input{plots/nex-hist}

Instead of storing two bytes for the number of exceptions we can bit pack the
number of exceptions using four bits to represent the number of bits used.
Refer back to Section \ref{sec:bitpack} for a review of bit packing.
The maximum number of exceptions, 3616, requires at least
12 bits to represent it. Using the bit packing approach, $4+12=16$ bits or two
bytes would be used to represent it -- which is equivalent to the naive
approach. That is, in the worst case bit packing uses the same space as the naive
approach. In the best case when there are zero exceptions only four bits are
required, and as long as there are less than 16 exceptions up to one byte is
used. Although this approach clearly saves space, we are talking about saving at
most 12 bits which is miniscule compared to the scale of the data.

Similarly, we can store the exceptions' data in a more compact form.
The exceptions themselves range from 256 to 2317 but 99\%
of the exceptions do not go higher than 481. The mean and mode are $\sim$277 and
256 respectively. Refer back to Table \ref{tab:ex} for a summary and Figure
\ref{fig:ex-hist} for the histogram. Notably, the frequency of even-numbered
exceptions is much greater than adjacent odd-numbered exceptions from 256 up to
$\sim$330. This is due to large positive deltas occuring more frequently than
large negative deltas as previously discussed in Section \ref{subsec:stripe}.
One simple approach to encoding the exceptions is to subtract 256 since this is
the minimum exception value and then apply bit packing using 4 bits as before. The
average exception can then be represented using
0 to 5 bits which is better than the naive 16-bit per exception encoding. This
results in an expected decrease of $(16-5)\times 5-4=51$ bits per read.
In the worst case, the number of bits used per exception for this data will be 12
or $4 + 12e$ in total since 2317 is the maximum exception. This is better than
the naive approach for $e>1$ and equivalent for $e=1$. The implications of this
is that bit packing the exceptions' data never consumes more space than the
naive strategy.

\input{plots/ex-hist}

Another improvement we can make is on the storage of the exceptions' positions.
Since the positions are strictly increasing we can take the deltas between each
position and subtract one. This will result in smaller values which we can then
bit pack using 5 bits instead of 4 for the number of bits used.
5 bits are required instead of 4 to account for any outliers given the maximum
read length is roughly 6 million. If 4 bits were used only position deltas up to
$2^{2^{4}-1}-1=32767$ could be represented which is clearly short of 6 million.

The strategy which combines all of the above exception bit packing strategies
and still stores the regular one byte data we will name \textit{compact vbbe21}.
Notice the extra `b' in \textit{vbbe21}. This refers to the bit packing
strategy in use.
See Figure \ref{fig:vbbe21-compact} for a visual representation of this
encoding. Notice the byte boundary padding which aligns the one byte data to
byte boundaries for downstream compression methods. For implementation
simplicity consider a similar strategy which does not bit pack the number of
exceptions, bit packs the exception' positions and data using one byte for the
number of bits used rather than 5 and 4 respectively, and uses padding between
the exceptions' positions and data. We shall name this strategy the
\textit{regular vbbe21} encoding or \textit{vbbe21} for short. See Figure
\ref{fig:vbbe21} for a pictorial representation.

\input{plots/vbbe21-compact}
\input{plots/vbbe21}

% TODO analyse how much better
The storage gap between vbbe21 and compact vbbe21 has an upper bound of
\begin{align*}
	\text{vbbe21} - \text{compact vbbe21} &= \text{number of exceptions gap}
	+ \text{positions gap}\\ &+\text{two byte exceptions gap} + \text{padding gap}\\
	&< (16-4) + (8-5) + (8-4) + 16\\
	&= 35
\end{align*}
bits
and a lower bound of
\[ \text{vbbe21} - \text{compact vbbe21} > (16-12) + (8-5) + (8-4) - 8 = 3 \]
bits.
That is, compact vbbe21 saves between 3 and 35 bits compared to regular vbbe21.
However, an expected measure of the storage gap is a more useful metric for
comparison. Let's name the non--one-byte-data where the exceptions are handled
the \textit{exceptions section}. Using Equation \ref{eq:vbe21}, the expected size of
the vbe21-zd exceptions section in bytes is given by
\[ 2 + 6E[X] = 36.02. \]
Now, round up the expected number of exceptions from Table \ref{tab:ex} so it is
5. Then, the expected number of bits used for the number of exceptions is roughly
3. Let the expected number of bits used on the positions be 16 since the expected
read length is \num{113471} and this accounts for position deltas up to
\num{65535}. Furthermore, let the expected number of bits used on the
exceptions be 5 since the average exception is 277 and this accounts for
exceptions up to 287. Also, let the expected padding size be 3.5 bits as this is
halfway on the padding's range of 0 to 7 bits.
Then, an estimate of the expected size of the vbbe21-zd exceptions section is
given by
\begin{align*}
	&\text{number of exceptions} + \text{positions} + \text{two byte exceptions} + \text{padding}\\
	&= \frac{16 + (8+5\times16) + (8+5\times 5) + 2\times3.5}{8}\\
	&= 18
\end{align*}
bytes, versus
\[ \frac{(4+3) + (5+5\times16) + (4+5\times 5) + 3.5}{8}= 15.56 \]
bytes
for the compact vbbe21-zd exceptions section. These sizes are displayed in Table
\ref{tab:ex-sec-exp}
for ease of comparison. Now, we can see that the estimated storage gap between
vbbe21-zd and compact vbbe21-zd is 2.44 bytes or 19.5 bits. Furthermore,
vbbe21-zd and compact vbbe21-zd are expected to save 17 and 19.44 bytes
per read respectively when compared to vbe21-zd. So, there is a storage benefit
to bit packing the exceptions over storing them in a raw sequential format.

\input{plots/ex-sec-exp}

% TODO compare improvement in results

In comparison with vbe21, the compact and regular vbbe21 algorithms still take
linear time during encoding and decoding although they require more computational
operations to encode and evaluate the bit packed data. However, the exceptions
section usually accounts for less than 0.05\% of the vbe21-zd encoding
compared to the one byte data. Hence, focussing
on compressing the one byte data section has much more potential benefit. This
will be heavily explored in the remaining sections.

